Taking Linear Logic Apart

### 2018-10-28 doc/paper/main.tex
final version

\begin{abstract}
  Process calculi based on logic, such as \piDILL and \cp, provide a foundation for deadlock-free concurrent programming. However, in previous work, there is a mismatch between the rules for constructing proofs and the term constructors of the \textpi-calculus. We introduce Hypersequent Classical Processes (\hcp), which addresses this mismatch using hypersequents (collections of sequents) to register parallelism in the typing judgements. We prove that \hcp enjoys deadlock-freedom and a series of properties that relate it back to \cp.
\end{abstract}

@@

In this paper, we introduce Hypersequent Classical Processes (\hcp), which addresses this mismatch. The key insight is to register parallelism in the typing judgements using hypersequents~\citep{avron1991}, a technique from logic which generalises judgements from one sequent to many. This allows us to take apart the term constructs used in Classical Processes (\cp) to more closely match those of the standard \textpi-calculus. We proceed as follows. We start by introducing \cp (\cref{sec:cp}). Then, we introduce \hcp and prove it enjoys subject reduction and progress (\cref{sec:hcp}). We prove that every \cp process is an \hcp process, and that \hcp supports the same communication protocols as \cp, and no more (\cref{sec:cp2hcp}). Finally, we discuss related work (\cref{sec:related-work}).

@@

The term language of \cp is a variant of the \textpi-calculus. The variables $\tm{x}$, $\tm{y}$, and $\tm{z}$ range over channel names. The construct $\tm{\cpLink{x}{y}}$ links two channels~\citep{sangiorgi1996,boreale1998}, forwarding messages received on $\tm{x}$ to $\tm{y}$ and vice versa. The construct $\tm{\cpCut{x}{P}{Q}}$ creates a new channel $\tm{x}$, and composes two processes, $\tm{P}$ and $\tm{Q}$, which communicate on $\tm{x}$, in parallel. Therefore, in $\tm{\cpCut{x}{P}{Q}}$ the name $\tm{x}$ is bound in both $\tm{P}$ and $\tm{Q}$. In $\tm{\cpRecv{x}{y}{P}}$ and $\tm{\cpSend{x}{y}{P}{Q}}$, round brackets denote input, square brackets denote output. \cp uses bound output~\citep{sangiorgi1996}, meaning that both input and output bind a new name. In $\tm{\cpRecv{x}{y}{P}}$ the new name $\tm{y}$ is bound in $\tm{P}$. In $\tm{\cpSend{x}{y}{P}{Q}}$, the new name $\tm{y}$ is only bound in $\tm{P}$, while $\tm{x}$ is only bound in $\tm{Q}$.
\begin{definition}[Terms]\label{def:cp-terms}
  Process terms are given by the following grammar:
      &  \mid& \tm{\cpCut{x}{P}{Q}}     &\text{parallel composition, ``cut''}
      \\ \mid& \tm{\cpSend{x}{y}{P}{Q}} &\text{output}
      &  \mid& \tm{\cpRecv{x}{y}{P}}    &\text{input}
      \\ \mid& \tm{\cpHalt{x}}          &\text{halt}
      &  \mid& \tm{\cpWait{x}{P}}       &\text{wait}
Terms in \cp are identified up to structural congruence, which states that links are symmetric, and parallel compositions $\tm{\cpCut{x}{P}{Q}}$ are associative and commutative.
\begin{definition}[Structural congruence]\label{def:cp-equiv}
  The structural congruence $\equiv$ is the congruence closure over terms which
  satisfies the following additional axioms:
      \cpEquivCutComm
      & \tm{\cpCut{x}{Q}{P}}
      \cpEquivCutAss1
      & \tm{\cpCut{x}{P}{\cpCut{y}{Q}{R}}}
      & \tm{\cpCut{y}{\cpCut{x}{P}{Q}}{R}}
      & \text{if }\notFreeIn{x}{R}\text{ and }\notFreeIn{y}{P}
Reductions relate processes with their reduced forms \eg a reduction $\reducesto{P}{Q}$ denotes that the process $\tm{P}$ can reduce to the process $\tm{Q}$ in a single step.
\begin{definition}[Reduction]\label{def:cp-reduction}
  Reductions are described by the smallest relation $\Longrightarrow$ on process terms closed under rules below.
        \cpRedAxCut1
        \cpRedBetaTensParr
        & \tm{\cpCut{x}{\cpSend{x}{y}{P}{Q}}{\cpRecv{x}{y}{R}}}
        & \tm{\cpCut{y}{P}{\cpCut{x}{Q}{R}}}
        \cpRedBetaOneBot
        & \tm{\cpCut{x}{\cpHalt{x}}{\cpWait{x}{P}}}
        \cpRedBetaPlusWith1
        \cpRedBetaPlusWith2
        \SYM{\cpRedGammaCut}
        \UIC{$\reducesto{\cpCut{x}{P}{Q}}{\cpCut{x}{P^\prime}{Q}}$}
        \SYM{\cpRedGammaEquiv}

@@

Duality plays a crucial role in both linear logic and session types. In \cp, the two endpoints of a channel are assigned dual types. This ensures that, for instance, whenever a process \emph{sends} across a channel, the process on the other end of that channel is waiting to \emph{receive}. Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$. Duality is an involutive function \ie $\ty{(A^\bot)^\bot} = \ty{A}$.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \setlength{\arraycolsep}{3pt}
    \begin{array}{lclclclclclclcl}
               \ty{(A \tens B)^\bot} & = & \ty{A^\bot \parr B^\bot}
      &\qquad& \ty{\one^\bot}        & = & \ty{\bot}
      &\qquad& \ty{(A \parr B)^\bot} & = & \ty{A^\bot \tens B^\bot}
      &\qquad& \ty{\bot^\bot}        & = & \ty{\one}
      \\       \ty{(A \plus B)^\bot} & = & \ty{A^\bot \with B^\bot}
      &\qquad& \ty{\nil^\bot}        & = & \ty{\top}
      &\qquad& \ty{(A \with B)^\bot} & = & \ty{A^\bot \plus B^\bot}
      &\qquad& \ty{\top^\bot}        & = & \ty{\nil}
    \end{array}
  \]
\end{definition}\noindent
An environment associates channels with types. Names in environments must be unique, and two environments $\ty{\Gamma}$ and $\ty{\Delta}$ can only be combined as $\ty{\Gamma, \Delta}$ if $\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} \Coloneqq \ty{\emptycontext} \mid \ty{\Gamma}, \tmty{x}{A}
  \)
\end{definition}\noindent
A typing judgement associates a process with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}, \dots, \tmty{x_n}{A_n}}$
  denotes that the process \tm{P} communicates along channels $\tm{x_1}$, \dots,
  $\tm{x_n}$ following protocols $\ty{A_1}$, \dots, $\ty{A_n}$. Typing
  judgements are derived using rules below.
  \\
  {Structural rules}
  \begin{center}
    \cpInfAx
    \cpInfCut
  \end{center}
	{Logical rules}
  \begin{center}
    \cpInfTens
    \cpInfParr
  \end{center}
  \begin{center}
    \cpInfBot
    \cpInfOne
  \end{center}
  \begin{center}
    \cpInfPlus1
    \cpInfPlus2
  \end{center}
  \begin{center}
    \cpInfWith
  \end{center}
  \begin{center}
    \cpInfNil
    \cpInfTop   
  \end{center}
\end{definition}

@@

In this section, we introduce Hypersequent Classical Processes (\hcp), a variant of \cp which registers parallelism in the typing judgements using hypersequents, allowing us to take apart the monolithic term constructors of \cp (\eg $\tm{\cpSend{x}{y}{P}{Q}}$) into the corresponding \textpi-calculus term constructs.

@@

\subsection{Types}
We use the same definitions for types and environments for \hcp as we used for \cp. However, we introduce a new layer on top of sequents: hypersequents. As \cp is a one-sided logic, and it uses the left-hand side of the turnstile to write the process, the traditional hypersequent notation can look confusing: ``$\seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}$'' seems to claim that $\tm{P}$ acts according to protocol $\ty{\Gamma_1}$. What are all the other $\Gamma$s doing there? Are they typing empty processes? Therefore, we opt to leave out the repeated turnstile, and instead work with the notion of ``hyper-environments''. However, we will still refer to our system as a hypersequent system. A hyper-environment is either empty, or consist of a series of typing environments, separated by vertical bars. A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ types a series of $n$ entangled, but independent processes.
\begin{definition}[Hyper-environments]\label{def:hcp-hyper-environment}
  $\ty{\mathcal{G}}, \ty{\mathcal{H}} \Coloneqq \ty{\emptyhypercontext} \, \mid \, \ty{\mathcal{G} \hsep \Gamma}$
\end{definition}\noindent
A hyper-environment is a multiset of environments. While names within environments must be unique, names may be shared between multiple environments in a hyper-environment. We write $\ty{\mathcal{G} \hsep \mathcal{H}}$ to combine two hyper-environments.

Typing judgements in \hcp associate processes with hyper-environments. \textsc{H-Mix} composes two processes in parallel, but remembers that they are independent in the sequent. \textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of at least two independent processes, and connects them, eliminating the vertical bar. Each logical rule has the side condition that $\tm{x} \not\in \ty{\mathcal{G}}$, which can be read as ``you cannot act on one end-point of $x$ if you are also holding its other end-point''. This prevents self-locking processes, \eg $\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes that the process $\tm{P}$ consists of $n$ independent, but potentially entangled processes, each of which communicates according to its own protocol $\Gamma_i$. 
  Typing judgements can be constructed using the inference rules below. 
  \\
  {Structural rules}
  \begin{center}
    \hcpInfAx
    \hcpInfCut
  \end{center}
  \begin{center}
    \hcpInfMix
    \hcpInfHalt
  \end{center}
  {Logical rules}
  \begin{center}
    \hcpInfBoundTens
    \hcpInfParr
  \end{center}
  \begin{center}
    \hcpInfOne
    \hcpInfBot
  \end{center}
  \begin{center}
    \hcpInfPlus1
    \hcpInfPlus2
  \end{center}
  \begin{center}
    \hcpInfWith
  \end{center}
  \begin{center}
    \hcpInfNil
    \hcpInfTop
  \end{center}
  Furthermore, each logical rule has the side condition that $\tm{x} \not\in \ty{\mathcal{G}}$.
\end{definition}\noindent

@@

In this section, we discuss the relationship between \cp and \hcp. We will prove two important theorems: every \cp process is an \hcp process; and \hcp supports the same protocols as \cp. We define a translation from terms in \cp to terms in \hcp which breaks down the term constructs in \cp into their more atomic constructs in \hcp.
\begin{definition}\label{def:cp2hcp-terms}
  \[
    \begin{array}{lcllcl}
         \tm{\mtf{\cpLink{x}{y}}}
      &  \coloneqq & \tm{\cpLink{x}{y}}
      &  \tm{\mtf{\cpCut{x}{P}{Q}}}
      &  \coloneqq & \tm{\piNew{x}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      \\ \tm{\mtf{\cpSend{x}{y}{P}{Q}}}
      &  \coloneqq & \tm{\piBoundSend{x}{y}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      &  \tm{\mtf{\cpRecv{x}{y}{P}}}
      &  \coloneqq & \tm{\piRecv{x}{y}{\mtf{P}}}
      \\ \tm{\mtf{\cpHalt{x}}}
      &  \coloneqq & \tm{\piBoundSend{x}{}{\piHalt}}
      &  \tm{\mtf{\cpWait{x}{P}}}
      &  \coloneqq & \tm{\piRecv{x}{}{\mtf{P}}}
      \\ \tm{\mtf{\cpInl{x}{P}}}
      &  \coloneqq & \tm{\cpInl{x}{\mtf{P}}}
      &  \tm{\mtf{\cpInr{x}{P}}}
      &  \coloneqq & \tm{\cpInr{x}{\mtf{P}}}
      \\ \tm{\mtf{\cpCase{x}{P}{Q}}}
      &  \coloneqq & \tm{\cpCase{x}{\mtf{P}}{\mtf{Q}}}
      &  \tm{\mtf{\cpAbsurd{x}}}
      &  \coloneqq & \tm{\cpAbsurd{x}}
    \end{array}
  \]
\end{definition}\noindent
We will use this relation in the first proof, and its analogue for derivations in the second.

@@

\subsection{\hcp supports the same communication protocols as \cp}
\hcp exhibits some behaviours which are impossible to directly translate back into \cp. For instance, in the process below, the choice sent on $\tm{x}$ will affect the choice between $\tm{P}$ and $\tm{P'}$, even though neither has access to the channel $\tm{x}$.
\[
\begin{prooftree*}
  \AXC{}\noLine\UIC{$\seq[P]{ \Gamma }$}
  \AXC{}\noLine\UIC{$\seq[Q]{ \Delta, \tmty{x}{A} }$}
  \NOM{H-Mix}
  \BIC{$\seq[\piPar{P}{Q}]{ \Gamma \hsep \Delta, \tmty{x}{A} }$}
  \AXC{}\noLine\UIC{$\seq[P']{ \Gamma }$}
  \AXC{}\noLine\UIC{$\seq[Q']{ \Delta, \tmty{x}{B} }$}
  \NOM{H-Mix}
  \BIC{$\seq[\piPar{P'}{Q'}]{ \Gamma \hsep \Delta, \tmty{x}{B} }$}
  \SYM{\with}
  \BIC{$\seq[\cpCase{x}{\piPar{P}{Q}}{\piPar{P'}{Q'}}]{ \Gamma \hsep \Delta, \tmty{x}{A \with B}}$}
\end{prooftree*}
\]
Instead, we will prove that \hcp supports the same communication protocols as \cp. This is the same as saying that it inhabits the same session types, or that the associated logical systems derive the same theorems. We show this by proving that we can internalise the hyper-environments as formulas in the logic. This is a standard method for proving the soundness of a hypersequent calculus.

@@

  Disentanglement is described by the smallest relation $\moveMixDown$ on proof derivations closed under the rules in \cref{fig:hcp-disentangle}, plus the associativity and commutativity of mixes. The relation $\moveMixDown^\star$ is the reflexive, transitive closure of $\moveMixDown$.

@@

  \(

@@

    \pvar{$\rho_1$}\UIC{$\seq{\mathcal{G}_1 \hsep {A}, \Gamma}$}
    \pvar{$\rho_2$}\UIC{$\seq{\mathcal{G}_2}$}

@@

    \BIC{$\seq{\mathcal{G}_1 \hsep \mathcal{G}_2}$}
    \pvar{$\rho_3$}\UIC{$\seq{\mathcal{G}_3 \hsep {A^\bot}, \Delta}$}

@@

    \BIC{$\seq{\mathcal{G}_1 \hsep \mathcal{G}_2 \hsep \mathcal{G}_3 \hsep \Gamma, \Delta}$}

@@

    & \pvar{$\rho_1$}\UIC{$\seq{\mathcal{G}_1 \hsep {A}, \Gamma}$}
      \pvar{$\rho_3$}\UIC{$\seq{\mathcal{G}_3 \hsep {A^\bot}, \Delta}$}

@@

      \BIC{$\seq{\mathcal{G}_1 \hsep \mathcal{G}_3 \hsep \Gamma, \Delta}$}
      \pvar{$\rho_2$}\UIC{$\seq{\mathcal{G}_2}$}

@@

      \BIC{$\seq{\mathcal{G}_1 \hsep \mathcal{G}_2 \hsep \mathcal{G}_3 \hsep \Gamma, \Delta}$}

@@

    \pvar{$\rho_1$}\UIC{$\seq{\mathcal{G}_1 \hsep {A}, \Gamma}$}
    \pvar{$\rho_2$}\UIC{$\seq{\mathcal{G}_2}$}

@@

    \BIC{$\seq{\mathcal{G}_1 \hsep \mathcal{G}_2 \hsep {A}, \Gamma}$}
    \pvar{$\rho_3$}\UIC{$\seq{\mathcal{G}_3 \hsep {B}, \Delta}$}

@@

    \BIC{$\seq{\mathcal{G}_1 \hsep \mathcal{G}_2 \hsep \mathcal{G}_3
    \hsep {A \tens B}, \Gamma, \Delta}$}

@@

  \)

>>>>

\begin{abstract}
% Process calculi based on logic, such as \piDILL and \cp, provide a foundation for deadlock-free concurrent programming. However, in previous work, there is a mismatch between the rules for constructing proofs and the term constructors of the \textpi-calculus. We introduce Hypersequent Classical Processes (\hcp), which addresses this mismatch using hypersequents (collections of sequents) to register parallelism in the typing judgements. We prove that \hcp enjoys deadlock-freedom and a series of properties that relate it back to \cp.
\end{abstract}

@@

\citet{kokke2019} introduced Hypersequent Classical Processes (\hcp), which addresses this mismatch. The key insight is to register parallelism in the typing judgements using hypersequents~\citep{avron1991}, a technique from logic which generalises judgements from one sequent to many. This allows us to take apart the term constructs used in Classical Processes (\cp) to more closely match those of the standard \textpi-calculus. \hcp has labelled transition semantics with delayed actions~\citep{merro2004} and a full abstraction result: bisimilarity, denotational equivalence, and barbed congruence coincide.

However, the step from \cp to \hcp is a big one. As of yet, \hcp does not have reduction semantics, and the addition of delayed actions means that \cp processes interpreted as \hcp processes do not behave as they would in \cp.

In this paper, we address these issues by introducing a variant of \hcp with reduction semantics and without delayed actions. We proceed as follows. We start by introducing \cp (\cref{sec:cp}). Then, we introduce our variant of \hcp and prove it enjoys subject reduction and progress (\cref{sec:hcp}). We prove that every \cp process is an \hcp process, relate processes in \hcp back to \cp, and restate the result that \hcp supports the same communication protocols as \cp (\cref{sec:cp2hcp}). Finally, we discuss related work (\cref{sec:related-work}).

@@

\subsection{Terms}
The term language of \cp is a variant of the \textpi-calculus. The variables $\tm{x}$, $\tm{y}$, and $\tm{z}$ range over channel names. The construct $\tm{\cpLink{x}{y}}$ links two channels~\citep{sangiorgi1996,boreale1998}, forwarding messages received on $\tm{x}$ to $\tm{y}$ and vice versa. The construct $\tm{\cpCut{x}{P}{Q}}$ creates a new channel $\tm{x}$, and composes two processes, $\tm{P}$ and $\tm{Q}$, which communicate on $\tm{x}$, in parallel. Therefore, in $\tm{\cpCut{x}{P}{Q}}$ the name $\tm{x}$ is bound in both $\tm{P}$ and $\tm{Q}$. In $\tm{\cpRecv{x}{y}{P}}$ and $\tm{\cpSend{x}{y}{P}{Q}}$, round brackets denote input, square brackets denote output. \cp uses bound output~\citep{sangiorgi1996}, meaning that both input and output bind a new name. In $\tm{\cpRecv{x}{y}{P}}$ the new name $\tm{y}$ is bound in $\tm{P}$. In $\tm{\cpSend{x}{y}{P}{Q}}$, the new name $\tm{y}$ is only bound in $\tm{P}$, while $\tm{x}$ is only bound in $\tm{Q}$.
\begin{definition}[Terms]\label{def:cp-terms}
  Process terms are given by the following grammar:
  \[
    \begin{array}{rllrll}
      \tm{P}, \tm{Q}, \tm{R}
           \Coloneqq & \tm{\cpLink{x}{y}}       &\text{link}
      &  \mid& \tm{\cpCut{x}{P}{Q}}     &\text{parallel composition, ``cut''}
      \\ \mid& \tm{\cpSend{x}{y}{P}{Q}} &\text{output}
      &  \mid& \tm{\cpRecv{x}{y}{P}}    &\text{input}
      \\ \mid& \tm{\cpHalt{x}}          &\text{halt}
      &  \mid& \tm{\cpWait{x}{P}}       &\text{wait}
      \\ \mid& \tm{\cpInl{x}{P}}        &\text{select left choice}
      &  \mid& \tm{\cpInr{x}{P}}        &\text{select right choice}
      \\ \mid& \tm{\cpCase{x}{P}{Q}}    &\text{offer binary choice}
      &  \mid& \tm{\cpAbsurd{x}}        &\text{offer nullary choice}
    \end{array}
  \]
\end{definition}\noindent
Terms in \cp are identified up to structural congruence, which states that links are symmetric, and parallel compositions $\tm{\cpCut{x}{P}{Q}}$ are associative and commutative.
\begin{definition}[Structural congruence]\label{def:cp-equiv}
  The structural congruence $\equiv$ is the congruence closure over terms which
  satisfies the following additional axioms:
  \[
    \setlength{\arraycolsep}{3pt}
    \begin{array}{llcll}
      \cpEquivLinkComm
      & \tm{\cpLink{x}{y}}
      & \equiv
      & \tm{\cpLink{y}{x}}
      \\
      \cpEquivCutComm
      & \tm{\cpCut{x}{P}{Q}}
      & \equiv
      & \tm{\cpCut{x}{Q}{P}}
      \\
      \cpEquivCutAss1
      & \tm{\cpCut{x}{P}{\cpCut{y}{Q}{R}}}
      & \equiv
      & \tm{\cpCut{y}{\cpCut{x}{P}{Q}}{R}}
      & \text{if }\notFreeIn{x}{R}\text{ and }\notFreeIn{y}{P}
    \end{array}
  \]
\end{definition}\noindent
The reduction semantics presented here are a variant of those presented by \citet{lindley2015}, who showed that reduction in \cp can be decomposed in two phases: one in which all \textbeta-reduction happens, and one in which an arbitrary action, blocked on an external communication, is moved to the top of the term using commuting conversions. We choose to stop after the first phase, and do away with the commuting conversions.

Reductions relate processes with their reduced forms \eg a reduction $\reducesto{P}{Q}$ denotes that the process $\tm{P}$ can reduce to the process $\tm{Q}$ in a single step.
\begin{definition}[Reduction]\label{def:cp-reduction}
  Reductions are described by the smallest relation $\Longrightarrow$ on process terms closed under rules below.
  \begin{gather*}
      \begin{array}{llcll}
        \cpRedAxCut1
        & \tm{\cpCut{x}{\cpLink{w}{x}}{P}}
        & \Longrightarrow
        & \tm{\cpSub{w}{x}{P}}
        \\
        \cpRedBetaTensParr
        & \tm{\cpCut{x}{\cpSend{x}{y}{P}{Q}}{\cpRecv{x}{y}{R}}}
        & \Longrightarrow
        & \tm{\cpCut{y}{P}{\cpCut{x}{Q}{R}}}
        \\
        \cpRedBetaOneBot
        & \tm{\cpCut{x}{\cpHalt{x}}{\cpWait{x}{P}}}
        & \Longrightarrow
        & \tm{P}
        \\
        \cpRedBetaPlusWith1
        & \tm{\cpCut{x}{\cpInl{x}{P}}{\cpCase{x}{Q}{R}}}
        & \Longrightarrow
        & \tm{\cpCut{x}{P}{Q}}
        \\
        \cpRedBetaPlusWith2
        & \tm{\cpCut{x}{\cpInr{x}{P}}{\cpCase{x}{Q}{R}}}
        & \Longrightarrow
        & \tm{\cpCut{x}{P}{R}}
      \end{array}
    \\
      \begin{prooftree*}
        \AXC{$\reducesto{P}{P^\prime}$}
        \SYM{\cpRedGammaCut}
        \UIC{$\reducesto{\cpCut{x}{P}{Q}}{\cpCut{x}{P^\prime}{Q}}$}
      \end{prooftree*}
      \qquad
      \begin{prooftree*}
        \AXC{$\tm{P}\equiv\tm{Q}$}
        \AXC{$\reducesto{Q}{Q^\prime}$}
        \AXC{$\tm{Q^\prime}\equiv\tm{P^\prime}$}
        \SYM{\cpRedGammaEquiv}
        \TIC{$\reducesto{P}{P^\prime}$}
      \end{prooftree*}
    \end{gather*}
  Relations $\Longrightarrow^{+}$ and $\Longrightarrow^\star$ are the transitive, and the reflexive, transitive closures of $\Longrightarrow$, respectively.
\end{definition}

@@

Duality plays a crucial role in both linear logic and session types. In \cp, the two endpoints of a channel are assigned dual types. This ensures that, for instance, whenever a process \emph{sends} across a channel, the process on the other end of that channel is waiting to \emph{receive}. Each type $\ty{A}$ has a dual, written $\ty{A^\bot}$. Duality is an involutive function \ie $\ty{(A^\bot)^\bot} = \ty{A}$.
\begin{definition}[Duality]\label{def:cp-negation}
  \[
    \setlength{\arraycolsep}{3pt}
    \begin{array}{lclclclclclclcl}
               \ty{(A \tens B)^\bot} & = & \ty{A^\bot \parr B^\bot}
      &\qquad& \ty{\one^\bot}        & = & \ty{\bot}
      &\qquad& \ty{(A \parr B)^\bot} & = & \ty{A^\bot \tens B^\bot}
      &\qquad& \ty{\bot^\bot}        & = & \ty{\one}
      \\       \ty{(A \plus B)^\bot} & = & \ty{A^\bot \with B^\bot}
      &\qquad& \ty{\nil^\bot}        & = & \ty{\top}
      &\qquad& \ty{(A \with B)^\bot} & = & \ty{A^\bot \plus B^\bot}
      &\qquad& \ty{\top^\bot}        & = & \ty{\nil}
    \end{array}
  \]
\end{definition}\noindent
An environment associates channels with types. Names in environments must be unique, and two environments $\ty{\Gamma}$ and $\ty{\Delta}$ can only be combined as $\ty{\Gamma, \Delta}$ if $\ty{\fv{\Gamma}} \cap \ty{\fv{\Delta}} = \varnothing$.
\begin{definition}[Environments]\label{def:cp-environments}
  \(
    \ty{\Gamma}, \ty{\Delta}, \ty{\Theta} \Coloneqq \ty{\emptycontext} \mid \ty{\Gamma}, \tmty{x}{A}
  \)
\end{definition}\noindent
A typing judgement associates a process with collections of typed channels.
\begin{definition}[Typing judgements]\label{def:cp}
  A typing judgement $\seq[{ P }]{\tmty{x_1}{A_1}, \dots, \tmty{x_n}{A_n}}$
  denotes that the process \tm{P} communicates along channels $\tm{x_1}$, \dots,
  $\tm{x_n}$ following protocols $\ty{A_1}$, \dots, $\ty{A_n}$. Typing
  judgements are derived using rules below.
  \\[1\baselineskip]
  {Structural rules}
  \begin{center}
    \cpInfAx \cpInfCut
  \end{center}
	{Logical rules}
  \begin{center}
    \cpInfTens \cpInfParr
  \end{center}
  \begin{center}
    \cpInfBot
    \cpInfOne
  \end{center}
  \begin{center}
    \cpInfPlus1
    \cpInfPlus2
  \end{center}
  \begin{center}
    \cpInfWith
  \end{center}
  \begin{center}
    \cpInfNil
    \cpInfTop   
  \end{center}
\end{definition}

@@

In this section, we introduce our variant of Hypersequent Classical Processes (\hcp), itself a variant of \cp which registers parallelism in the typing judgements using hypersequents, allowing us to take apart the monolithic term constructors of \cp (\eg $\tm{\cpSend{x}{y}{P}{Q}}$) into the corresponding \textpi-calculus term constructs.

The crucial difference between \hcp as described here and \hcp as described by \citet{kokke2019} is in the absence of delayed actions. However, removing delayed actions introduces self-locking processes, which we rule out using an extra restriction in the type system (see \cref{sec:hcp-types}). Furthermore, we follow \cp in using the same name for both endpoints of a channel, writing, \eg $\tm{\cpCut{x}{\cpHalt{x}}{\cpWait{x}{P}}}$ as opposed to $\tm{\cpCut{xy}{\cpHalt{x}}{\cpWait{y}{P}}}$.

@@

\subsection{Types}\label{sec:hcp-types}
We use the same definitions for types and environments for \hcp as we used for \cp. However, we introduce a new layer on top of sequents: hypersequents. As \cp is a one-sided logic, and it uses the left-hand side of the turnstile to write the process, the traditional hypersequent notation can look confusing: ``$\seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}$'' seems to claim that $\tm{P}$ acts according to protocol $\ty{\Gamma_1}$. What are all the other $\Gamma$s doing there? Are they typing empty processes? Therefore, we opt to leave out the repeated turnstile, and instead work with the notion of ``hyper-environments''. However, we will still refer to our system as a hypersequent system. A hyper-environment is either empty, or consist of a series of typing environments, separated by vertical bars. A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ types a series of $n$ entangled, but independent processes.
\begin{definition}[Hyper-environments]\label{def:hcp-hyper-environment}
  $\ty{\mathcal{G}}, \ty{\mathcal{H}} \Coloneqq \ty{\emptyhypercontext} \, \mid \, \ty{\mathcal{G} \hsep \Gamma}$
\end{definition}\noindent
A hyper-environment is a multiset of environments. While names within environments must be unique, names may be shared between multiple environments in a hyper-environment. We write $\ty{\mathcal{G} \hsep \mathcal{H}}$ to combine two hyper-environments.

Typing judgements in \hcp associate processes with hyper-environments. \textsc{H-Mix} composes two processes in parallel, but remembers that they are independent in the sequent. \textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of at least two independent processes, and connects them, eliminating the vertical bar. Each logical rule has the side condition that $\tm{x} \not\in \ty{\mathcal{G}}$, which can be read as ``you cannot act on one end-point of $x$ if you are also holding its other end-point''. This prevents self-locking processes, \eg $\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes that the process $\tm{P}$ consists of $n$ independent, but potentially entangled processes, each of which communicates according to its own protocol $\Gamma_i$. 
  Typing judgements can be constructed using the inference rules below. 
  \\[1\baselineskip]
  {Structural rules}
  \begin{center}
    \hcpInfAx
    \hcpInfCut
  \end{center}
  \begin{center}
    \hcpInfMix
    \hcpInfHalt
  \end{center}
  {Logical rules}
  \begin{center}
    \hcpInfBoundTens
    \hcpInfParr
  \end{center}
  \begin{center}
    \hcpInfOne
    \hcpInfBot
  \end{center}
  \begin{center}
    \hcpInfPlus1
    \hcpInfPlus2
  \end{center}
  \begin{center}
    \hcpInfWith
  \end{center}
  \begin{center}
    \hcpInfNil
    \hcpInfTop
  \end{center}
  Furthermore, each logical rule has the side condition that $\tm{x} \not\in \ty{\mathcal{G}}$.
\end{definition}\noindent

@@

In this section, we discuss the relationship between \cp and \hcp. We prove two important theorems: every \cp process is an \hcp process; and \hcp supports the same protocols as \cp. We define a translation from terms in \cp to terms in \hcp which breaks down the term constructs in \cp into their more atomic constructs in \hcp.
\begin{definition}\label{def:cp2hcp-terms}
  \[
    \begin{array}{lcllcl}
         \tm{\mtf{\cpLink{x}{y}}}
      &  \coloneqq & \tm{\cpLink{x}{y}}
      &  \tm{\mtf{\cpCut{x}{P}{Q}}}
      &  \coloneqq & \tm{\piNew{x}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      \\ \tm{\mtf{\cpSend{x}{y}{P}{Q}}}
      &  \coloneqq & \tm{\piBoundSend{x}{y}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      &  \tm{\mtf{\cpRecv{x}{y}{P}}}
      &  \coloneqq & \tm{\piRecv{x}{y}{\mtf{P}}}
      \\ \tm{\mtf{\cpHalt{x}}}
      &  \coloneqq & \tm{\piBoundSend{x}{}{\piHalt}}
      &  \tm{\mtf{\cpWait{x}{P}}}
      &  \coloneqq & \tm{\piRecv{x}{}{\mtf{P}}}
      \\ \tm{\mtf{\cpInl{x}{P}}}
      &  \coloneqq & \tm{\cpInl{x}{\mtf{P}}}
      &  \tm{\mtf{\cpInr{x}{P}}}
      &  \coloneqq & \tm{\cpInr{x}{\mtf{P}}}
      \\ \tm{\mtf{\cpCase{x}{P}{Q}}}
      &  \coloneqq & \tm{\cpCase{x}{\mtf{P}}{\mtf{Q}}}
      &  \tm{\mtf{\cpAbsurd{x}}}
      &  \coloneqq & \tm{\cpAbsurd{x}}
    \end{array}
  \]
\end{definition}\noindent
We use this relation in the first proof, and its analogue for derivations in the second.

@@

\subsection{\hcp supports the same communication protocols as \cp}
In this section, we prove that \hcp supports the same communication protocols as \cp. This is the same as saying that it inhabits the same session types, or that the associated logical systems derive the same theorems. We show this by proving that we can internalise the hyper-environments as formulas in the logic. This is a standard method for proving the soundness of a hypersequent calculus.

@@

  Disentanglement is described by the smallest relation $\moveMixDown$ on processes closed under the rules in \cref{fig:hcp-disentangle}, plus the associativity and commutativity of mixes. The relation $\moveMixDown^\star$ is the reflexive, transitive closure of $\moveMixDown$.

@@

@@

    \AXC{$\seq[P]{\mathcal{G} \hsep \Gamma, \tmty{x}{A} \hsep \Delta, \tmty{x}{A^\bot}}$}
    \AXC{$\seq[Q]{\mathcal{H}}$}

@@

    \BIC{$\seq[(P \ppar Q)]{\mathcal{G} \hsep \mathcal{H} \hsep \Gamma, \tmty{x}{A} \hsep \Delta, \tmty{x}{A^\bot}}$}

@@

    \UIC{$\seq[\piNew{x}{(P \ppar Q)}]{\mathcal{G} \hsep \mathcal{H} \hsep \Gamma, \Delta}$}

@@

    & \AXC{$\seq[P]{\mathcal{G} \hsep \Gamma, \tmty{x}{A} \hsep \Delta, \tmty{x}{A^\bot}}$}

@@

      \UIC{$\seq[\piNew{x}{P}]{\mathcal{G} \hsep \Gamma, \Delta}$}
      \AXC{$\seq[Q]{\mathcal{H}}$}

@@

      \BIC{$\seq[(\piNew{x}{P} \ppar Q)]{\mathcal{G} \hsep \mathcal{H} \hsep \Gamma, \Delta}$}

@@

    \AXC{$\seq[P]{\mathcal{G} \hsep \Gamma, \tmty{y}{A} \hsep \Delta, \tmty{x}{B}}$}
    \AXC{$\seq[Q]{\mathcal{H}}$}

@@

    \BIC{$\seq[(P \ppar Q)]{\mathcal{G} \hsep \mathcal{H} \hsep \Gamma, \tmty{y}{A} \hsep \Delta, \tmty{x}{B}}$}

@@

    \UIC{$\seq[\piBoundSend{x}{y}{(P \ppar Q)}]{\mathcal{G} \hsep \mathcal{H} \hsep \Gamma, \Delta, \tmty{x}{A \tens B}}$}

@@
